---
title: "Understanding MIRT/IRT"
author: "Adam Kurth"
date: "2024-01-18"
output: pdf_document
---

```{r setup, include=FALSE, warning=FALSE}
# install.packages("mirt")
library(mirt)
```

## Understanding MIRT/IRT

Will be rewriting over varaible names, for uniformity. 

### Given Simulation Code

```{r}
# trait level e.g. math skill
theta <- rnorm(n=1000, mean=0, sd=1)

# Item parameters (a, b, c) for 4 items
a <- c(0.5, 1, 1.5, 2)
b <- c(0, -1, 1, 0)
c <- c(0.2, 0.15, 0.25, 0.2)
```


```{r}
# for simulations with Monte Carlo experiments
n.persons = length(theta)
n.items = length(a)
response.data = matrix(NA, n.persons, n.items) # create empty 

for(i in 1:n.persons){
  for(j in 1:n.items){
    p <- c[j] + (1 -  c[j])/ (1 + exp(-(a[j]*theta[i] -b[j])))
    u <- runif(n=1, min = 0, max = 1)
    if(u < p){
      response.data[i,j] <- 1
    } else {
      response.data[i,j] <- 0
    }
  }
}

colnames(response.data) = c("I1", "I2", "I3", "I4")
```

calibrate item parameters using `mirt` package, default settings. 

```{r}
mirt.out = mirt::mirt(data= response.data, model = 1, itemtype = "3PL")
```

- `mirt.out`: results from mirt analysis.
- `data = response.data`: data matrix that was generated.
- `model = 1`: refers to the model should have 1 dimension.
- `itemtype = '3PL'` specifies that the item in the test should be a 3 parameter logistic model. 

```{r}
mirt::coef(mirt.out, simplify=T, IRTparts=T)
```
